The impact of earthquake cycle variability on neotectonic and paleoseismic slip rate estimates

### 2018-11-03 manuscript/reviews/response_letter.md

2. I don't think that we have a great understanding of the real mechanisms, 
   yet. There are a variety of mechanisms under consideration (e.g., co- and 
   post-seismic elastic and viscoelastic Coulomb stress changes, stress 
   transients, dynamic triggering, pore fluid pressure fluctuations, 
   fluctuations in the frictional failure threshold on a fault) in addition to 
   actual secular changes in tectonic loading rates. The time-dependent 
   mechanisms (particularly post-seismic processes) often show different 
   behavior with regard to whether they are 'spun-up' and at a dynamic 
   equilibrium, or not. And all of these mechanisms are necessarily linked to 
   uncertainty as to how (and where) faults are loaded to begin with--whether 
   the loading is in the elastic crust, in the viscoelastic/viscous mid or 
   lower crust (in a continuum style), or on a discrete creeping dislocation 
   down-dip of the brittle fault. There is a big range of scientific opinion on 
   all of these questions. As a community we are begging for a big review paper 
   to at least concatenate and organize these ideas and potentially test them 
   or at least sort them into compatible vs. mutually exclusive sets for future 
   testing. But we don't have that right now, so the topic is kind of a big 
   mudhole. I will dip my toe in but I really want to avoid falling in for the 
   purposes of this paper.

*Changes: New experiment with empirical slip distribution, and discussion of 
physical mechanisms behind aleatory recurrence variability.*

> [O5] There is a literature of ideas to draw upon, such as post-seismic fault 
> reloading (Kenner and Simons, 2005), earthquake super cycles (Sieh et al., 
> 2008; Weldon et al., 2004), isolated versus fault-network behavior (Berryman 
> et al., 2012). Some of these ideas are discussed briefly but need more 
> explanation.

I have added a short discussion (two paragraphs) on the topic, but as noted in 
my response to [O4] I don't think that a more full discussion is warranted. I 
want this paper to be a simple, easy-to-digest paper and I think that a long 
and necessarily unsatisfying discussion (as we don't have answers yet) on the 
mechanisms behind recurrence variability will be an obstacle, and many readers 
will just put the paper down.

@@

No, this isn't true. The outcome that I describe is a correlation between 
loading time and displacement, not an invariance of either loading time or 
displacement. Having 'the same' stress drop doesn't predict correlated loading 
time/recurrence intervals and displacement.

It's possible that this paragraph was too confusingly written for it to be 
easily interpretable. I have made some minor changes to the sentence structure 
for clarification

*Changes: Paragraph edits.*

## Review by Anonymous #2

> [B1] Uncertainty in estimates of earthquake recurrence and fault slip rate are
> important parameters, pursued by conscientious investigators of seismic
> hazard. The author perceives a lack of statistical support, and offers in
> this paper “insights” in variously tones of “friend of the practitioner” and
> “trust me, I’m the numerate one here”. Neither is convincing.

Someone's grumpy!

Since this review seems to aim to shame rather than help, the number of changes
are few.

> [B2a] A couple of omissions in this paper are particularly striking.

> [B2b] First, how do we have a paper addressed to “variability” in fault
> slip rate, addressing particularly the problem of small samples, without
> mentioning the methods of estimation for censored samples? There is an
> extensive statistical literature to estimate parameters and uncertainties and
> numerous recent papers applying it in paleoseismic contexts. This literature
> provides real quantitative methods to deal with the open intervals, long or
> short, that affect the geologist’s estimate of fault slip rate and recurrence
> estimation. These are real equations, with real uncertainties. One would look
> in vain in this paper for anything of similar substance.

Open intervals and censoring are of course well trodden ground in the
application of statistics to any field.

However, much of the variability that this paper is concerned with is present
within the closed intervals themselves. Open intervals

I don't disagree that more substance is offered by methods developed through the
decades by many researchers than I can offer in one short paper. Sorry!

> [B2c] Second, pages of this paper could be replaced (and improved) by a
> presentation and discussion of the properties of the standard error. E.g.,
> given an estimate a sample- based estimate of the mean, how far might the
> population (or true) mean be from the estimate? S.E. is estimated by the
> sample standard deviation divided by the square root of the number of
> samples. So, of course, estimates from small samples from a fuzzy log normal
> converge more slowly than from a well-defined (quasi-periodic) lognormal.
> Instead of a small equation (SE=s/sqrt(n)), our paper back-calculates the
> result using 2 million years of samples, and presents the results like a new
> discovery . And again, with little by way of meaningful uncertainties (e.g.,
> p1, lines 11- 13, 14- 16).

The standard error of the mean is irrelevant because slip rates are not
generally derived by taking the mean of independently-measured per-earthquake
offsets and recurrence intervals. They are typically derived by dividing net
offset over net time (including or excluding open intervals as one wishes in
paleoseismology, but never in neotectonics). Furthermore, with neotectonic slip
rate estimates, $n$ is not known.

It would be different if this paper was at all concerned with the calculation of
recurrence intervals, per-event displacements, or other parameters of individual
earthquakes where a number of samples

> [B3] 1 L 9,10: We read that the most important parameter is the coefficient of
>  variation. First, this equation is the arithmetic coefficient of variation,
>  and not the CV for a lognormal distribution. The CV of a lognormal does not
>  depend on the mean.

The reviewer is really confused about the lognormal distribution (see [B6] and
[B7] where the reviewer continues to make mistakes and level false accusations
at me).

I was quite worried that I made a mistake after reading this, so I re-checked
my math, my code and re-read a range of references (e.g., *Statistical
Distributions* by Evans, Hastings and Peacock, 1993; *Confidence Intervals
for the Coefficent of Variation for the Normal and Lognormal Distributions*,
Koopmans, Owen, and Rosenblatt, *Biometrika*, v. 51, no. 1/2, 1964, as well
as a lot of internet pages). Nowhere do I find alternative definitions of the
$CV$ for a lognormal distribution, except in the Wikipedia page for
Coefficient of Variation where two conflicting definitions for the Geometric
Coefficient of Variation are given. Both of these involve estimating the $CV$
from samples rather than as an algebraic transformation of the equations for
a lognormal distribution. The first definition is mathematically the same as
the standard definition; the difference lies in whether the estimation
procedure is done on the raw data or log-transformed data. The second is
simply the square of the first.

Just to reiterate, the arithmetic coefficient of variation for the lognormal
distribution is the standard definition.

The reviewer's confusion may stem from the fact that there are two equations
for the coefficient of variation that get regularly passed around. However, as I
demonstrate in the response to [B7c], they are mathematically equivalent. The
simple definition, which I use in the paper, is $CV = \frac{\sigma}{\mu}$,
where $\sigma$ and $\mu$ are the standard deviation and mean of the
distribution. The other definition, $CV=\sqrt{e^{\phi^2}-1}$, is given in
terms of the standard deviation $\phi$ of the natural log of the lognormal
distribution.

OK, this is pretty funny: All of the confusion that the reviewer has about the
lognormal distribution appears to be because of a misreading of the
Wikipedia pages for the lognormal distribution and for the coefficient of
variation. I mentioned the geometric CV (non-)issue above.  The quip that the
'CV of a lognormal does not depend on the mean' is... kind of true? As demonstrated in the response to [B7c], the standard deviation is a
*function of the mean* in that the mean is a coefficient of the standard
deviation. Therefore $CV = \sigma / \mu = \mu \Alpha /\mu$ where $\Alpha$ is
another mathematical expression (see [B7c]). So, yes, technically it is true but
only because it's on the top and the bottom of the fraction; furthermore, you
can't derive the expression $A$ without knowing both $\mu$ and $\sigma$.
Regardless, the quip gives all outward appearance of being a slight modification
of the sentence "Contrary to the arithmetic standard deviation, the arithmetic
coefficient of variation is independent of the arithmetic mean" which is below
the equation for the $CV$ of a lognormal on the Wikipedia page.  It is not a
statement that makes any practical sense unless one is beginning with
log-transformed data or distributions.

So to conclude, my math, code and analysis are sound in terms of the use of the
$CV$.

*No changes.*

> [B4] We could stop here, but a central flaw in the paper is exposed – nothing
> in this paper addresses how to obtain this most important parameter. If
> attempted, the essential emptiness of a 2,000,000 year sample would emerge. No
> real data set in paleoseismology resolves the mean and standard deviation to
> better than maybe 50%. Typical sites do well to resolve it to a factor of 2.
> P.3, line 22- 23 reflect this reality.

"the essential emptiness"... I'm surprised, and perhaps a bit concerned, to
find that I have tapped into the existential despair of Anonymous Reviewer #2.

This study investigates how slip rates measured as *offset / time* of an offset
marker are influenced by earthquake cycle variation. It is not a paper that
discusses the finer points of calculating the mean and standard deviation of
earthquake recurrence intervals or displacements from paleoseismology. Indeed,
given that "There is an extensive literature..." [B2b] on this topic, why should
I rehash it?

I fully agree that a typical paleoseismological study cannot resolve the mean
and standard deviation with a high degree of accuracy and I never stated
otherwise. Partly for this reason, by using CVs of 0.5, 1, and 2, I cover a
pretty wide range of values that can give some idea of the expected behavior of
the system under

> [B5a] It is not  obvious that the author has material experience words
> “aleatoric” and “epistemic”. Line 1, “aleatoric uncertainty” is a
> contradiction in terms.

It's really not a contradiciton in terms. 'Aleatory' does not mean 'the absence
of uncertainty', which it would have to in order to contradict the term
'uncertainty'.

Nonetheless

O'Hagan, T. (2004). Dicing with the unknown. *Significance*, 1(3), 132-133.

> [B5b] Bird, Zechar and Frankel all know better than to use the method the
> author alleges in lines 21 and 22 to arrive at epistemic uncertainty in slip
> rate.  They would more likely consider the allegation a misreading of their
> work.

This is a very perplexing comment. The line in question states, "The
uncertainty in the resulting slip estimate is typically treated as epistemic,
and quantified through the propagation of the measurement uncertainties on
the offset and time quantities (e.g., Bird, 2007; Zechar and Frankel, 2009)."
Both of the papers referenced are focused on estimating fault slip rates (or
*long-term offset rates* in Bird's terminology) that incorporate the measurement
uncertainty in both age and offset measurements. Both of their methods are based
in a convolution of the probability distributions for age and offset of the
marker features (i.e., the data with their uncertainties).

> [B5c] More broadly, the lack of care in writing makes one wonder how to
> understand this paper.

Sorry!

> [B5d] p2, L5:  A perturbation in slip rate would mean it was
> slipping at rate X, then changes to Y.

Yes.  This is both the intended and the obvious interpretation of the sentence.

*No changes.*

> [B5e] p.2, L14-17 have careful paleoseismologists doing reasonable things in
> one sentence, then imply they would make plainly rookie mistakes in the next.

The paragraph in question states that the open interval since the last
earthquake is often taken into consideration, however the variability in closed
intervals is not nearly as often taken into consideration--this is why I have
written the paper. Should I not state that paleoseismologists and
neotectonicists who consider the effects of the terminal open interval are
careful? I don't feel like being less generous. Should I not point out that
variability within the closed intervals is important and rarely considered?  No,
I think that I should.

*No changes.*

> [B5f] From here these read like inexperienced generalizations.

Sorry!

*No changes.*

> [B6a] p.4, L15-23: The descriptions of the lognormal variables here give one
> pause. First, log-normal parameters do not have units.

This comment is completely off the mark. The shape and scale parameters of a
lognormal distribution are indeed unitless, as they are the mean and standard
deviation of $\ln L$, where $L$ is a lognormal distribution (this transforms the
lognormal distribution into a normal distribution). However, the mean and
standard deviation of the actual lognormal distribution $L$ are in the units of
the distribution itself.  Nowhere in the manuscript did I state that $\mu$ and
$\sigma$, which are clearly defined as the mean and standard deviation of any
distribution in question, are the shape and scale parameters of $L$.  They are
simply the mean and standard deviation of it.

I have added a footnote clarifying this and giving the derivation of the shape
and scale parameters from the mean and standard deviation.

*Changes: Footnote with explanation and equations added.*

> [B7b] Second, the mean recurrence interval is not the location parameter of
> a log normal. This is just wrong.

Nowhere in the manuscript did I state this. In fact, the word 'location' is not
present anywhere in the submitted manuscript (it has been added to an expanded
discussion of probability distributions and their parameters in the revised
manuscript as described in the response to Comment [O14], though I still do not
make the mistake that I have been accused of).

Furthermore, the lognormal distribution doesn't have a location parameter--it is
a 2-parameter distribution specified by shape and scale parameters, as stated in
P. 3, L. 24 of the submitted manuscript.

*No changes.*

> [B7c] Third, if one uses the CV equation for the lognormal distribution (e.g.,
> https://en.wikipedia.org/wiki/Log-normal_distribution#Arithmetic_coefficient_of_variation),
> the CV will not match the COV alleged here. Given that the study depends on
> these distributions, we can’t really use subsequent conclusions.

This is false. The equations are mathematically equivalent.

The mean $\mu$ of a lognormal distribution $L$ is often given in terms of the
mean $m$ and standard deviation $\phi$ of the log-transformed distribution 
$\ln L$, which is a normal distribution. The equation is:

$$\mu = e^{m + \frac{1}{2} \phi^2 }$$

and similarly, the standard deviation $\sigma$ of $L$ may be given in terms of
$m$ and $\phi$:

$$\sigma = e^{m + \frac{1}{2} \phi^2} \sqrt{ e^{\phi^2} -1}\,,$$

which is clearly equivalent to

$$ \sigma = \mu \sqrt{ e^{\phi^2}-1}\,.$$

Therefore,

$$ CV = \frac{\sigma}{\mu} = \frac{\mu \sqrt{e^{\phi^2}-1}}{\mu} =
\sqrt{e^{\phi^2}-1}$$

which is the equation for the $CV$ of a lognormal distribution given in any
source including the Wikipedia page.  Please note that in most sources, the
symbols '$\mu$' and $\sigma$' are what I have here defined as '$m$' and '$\phi$'
because I use the symbols $\mu$ and $\sigma$ to represent the mean and standard
deviation of $L$, consistent with the manuscript where the same symbols are used
equivalently across distributions.

The methods and results of this study are consistent and this alleged error did
not occur.

*No changes.*

> [B7] p.6, L20.  If the number of samples is really n = N – t + 1, the samples
> are correlated by virtue of the overlap in the windows.  No accounting has
> been made of the correlation structure.

There is certainly a lot of autocorrelation of the samples if one were to
look at them sequentially. Nonetheless, even if sequential samples are
correlated, given enough samples, the final distribution of the samples won't
depend on the autocorrelation in the samples (this principle underlies the
family of powerful and commonly-used numerical methods based on Markov Chain
Monte Carlo sampling, for instance). This is why I use 2,000,000 years of
simulation: it ensures that the sampling is ample enough to not worry that
the serially-correlated samples have not sufficiently sampled the
distribution. I would have to use a much longer simulation (tens to hundreds
of millions of years) otherwise.

> [B8a] p.6, L26:  Starts a narrative of the consequences of the standard error,
> as though the standard error was never invented.

Again, the standard error is not applicable here because slip rates estimations
as used in this study are not made by taking the mean and standard deviation of
sets of measurements:  As laid out in

> [B8b] The fuzzy, back-of-the-envelope estimates start to get thick here. Real
> uncertainty estimates would serve better.

This is kept fuzzy because it's not possible to know the recurrence or offset distribtions

> [B9a] p.7, L1-3: Two observations: First, as written, the practicing
> geologist is being asked to believe that 60 earthquake cycles have passed
> with zero displacement. I can guess what was intended, but should not have to.

Actually, the naive reading of the sentence, that 60 *mean* earthquake cycles
can occur without displacement on a fault, is exactly what is intended.

The practicing geologist is probably familiar with intraplate faults in
environments with low strain rates.

> [B9b] Second, what probability is associated with this 60-cycle thing?  I ask
> because practicing hazard geologists have to make estimates, and give weights
> to extreme events.  What is the probability of 60 cycles, a CV of 2.0, ...?
> Hard to imagine that the author  has thought much about what these results
> would mean or how to use them if they were true.

I happen to be a practicing hazard geologist who is responsible for the
implementation of fault sources for probabilistic seismic hazard analysis, so
I have actually considered this

There are two probabilities here:
1. The probability of a fault with a CV of 2.0 going 60 mean earthquake cycles
   without accruing much displacement. This one is pretty easy to answer just

2. The probability of a fault having a CV of 2.0.  This is a goodFault 1 has a
   slip rate of 1 mm/yr, and Fault 2 has a slip rate of 0.1 mm/yr. If I think of
   nt of data.

So what should the practicing geologist do here? Well, that all depends on the
assumptions involved in the creation of the seismic source model. The first
thing to do i  The first thing 
to do i

### 2018-11-03 manuscript/eq_cycle_slip_rates.tex
updates to response following third review

No consensus exists among earthquake scientists as to the most
appropriate recurrence interval distribution. As is generally the case
with propriety, the safest and probably most correct assumption is that
it is context-dependent. Many studies of plate boundary faults such as
the San Andreas conclude that major or `characteristic' earthquakes are
periodic\\
\citep[e.g.,][]{berryman_major_2012,scharer_quasi-periodic_2010}.
Conversely, many intraplate faults with low slip rates appear to show
clustered earthquakes separated by long intervals of seismic quiescence
\citep[e.g.,][]{clark_long-term_2012}. However, one can find examples of
studies indicating the opposite conclusions
\citep{tuttle_earthquake_2002,grant_paleoseismic_1994}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  A \emph{periodic} distribution, represented by a lognormal
  distribution with a mean recurrence interval \(\mu\) = 1000 years, and
  a standard deviation \(\sigma\) = 500 years, and a \(CV\) = 0.5.
\item
  An \emph{unclustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 1000 years, and a \(CV\) =
  1.0.
\item
  An \emph{clustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 2000 years, and a \(CV\) =
  2.0.
\item
  An \emph{unclustered} time-independent distribution, represented by an
  exponential distribution with a mean recurrence interval \(\mu\) =
  1000 years, a standard deviation \(\sigma\) = 1000 years, and a
  \(CV\) = 1.0.
\end{enumerate}

>>>>

No consensus exists among earthquake scientists as to the most
appropriate recurrence interval distribution. As is generally the case
with propriety, the safest and probably most correct assumption is that
it is context-dependent. Many studies of plate boundary faults such as
the San Andreas conclude that major or `characteristic' earthquakes are
periodic\\
\citep[e.g.,][]{berryman_major_2012, scharer_quasi-periodic_2010}.
Conversely, many intraplate faults with low slip rates appear to show
clustered earthquakes separated by long intervals of seismic quiescence
\citep[e.g.,][]{clark_long-term_2012}. However, one can find examples of
studies indicating the opposite conclusions
\citep{tuttle_earthquake_2002, grant_paleoseismic_1994}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  A \emph{periodic} distribution, represented by a lognormal
  distribution with a mean recurrence interval \(\mu\) = 1000 years, and
  a standard deviation \(\sigma\) = 500 years, and a \(CV\) =
  0.5 \footnote{Please note that $\mu$ and $\sigma$ do not represent the scale
  and shape parameters of the lognormal distribution, though these symbols are
  commonly used to represent these parameters, as the scale and shape parameters
  of a lognormal distribution $L$ are the standard deviation and mean of the
  normal distribution $\ln L$. Instead, the shape parameter $sh$
  and scale parameter $sc$ are derived from $\mu$ and $\sigma$ through the
  relations $sh = \sqrt{1 + \sigma^2 / \mu^2}$ and 
  $sc = \ln [ \mu / \sqrt{1 + \sigma^2 / \mu^2}]$.}
\item
  An \emph{unclustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 1000 years, and a \(CV\) =
  1.0.
\item
  An \emph{clustered} time-dependent distribution, represented by a
  lognormal distribution with a mean recurrence interval \(\mu\) = 1000
  years, a standard deviation \(\sigma\) = 2000 years, and a \(CV\) =
  2.0.
\item
  An \emph{unclustered} time-independent distribution, represented by an
  exponential distribution with a mean recurrence interval \(\mu\) =
  1000 years, a standard deviation \(\sigma\) = 1000 years, and a
  \(CV\) = 1.0.
\end{enumerate}
